---
title: "Group Project"
output: html_document
---

##The code below properly formats the msaCounty Data, which will be needed to add MSA codes to the other tables worked with:
```{r}
#Reads in the downloaded datasets:
cancer <- read.csv(file='/Users/Tom/R/Group_Project/Data/lung_cancer_2016.csv',sep=',',header=T)
pollution <- read.csv(file='/Users/Tom/R/Group_Project/Data/Pollution_annual_all_2014.csv',sep=',',header=T)
msaCounty <- read.csv(file='/Users/Tom/R/Group_Project/Data/MSAandCounty.csv',sep=',',header=F,stringsAsFactors =FALSE)
for(j in 1:nrow(msaCounty)){
    msaCounty$V6[j] <- as.integer(msaCounty$V6[j])
}

#Makes integer values for county codes:
for(j in 1:nrow(msaCounty)){
    msaCounty$V4[j] <- as.integer(msaCounty$V4[j])
}

#Makes integer values for MSA codes:
for(j in 1:nrow(msaCounty)){
  if (is.na(msaCounty$V4[j])){
    msaCounty$V4[j] <- 0
  }
}

#Removes NA values and replaces themw with 0:
for(j in 1:nrow(msaCounty)){
  if (is.na(msaCounty$V6[j])){
    msaCounty$V6[j] <- 0
  }
}

#Replaces blank state names and county names with 0's:
for(z in 1:nrow(msaCounty)){
  if (msaCounty$V1[z]==""){
    msaCounty$V1[z] <- "ZZ"
  }
    if (msaCounty$V6[z]==""){
    msaCounty$V6[z] <- "ZZ"
  }
}

#Adds state codes to MSACounty table:
msaCounty$StateCode <-0
StateCode <- read.csv(file='/Users/Tom/R/Group_Project/Data/national_county.csv',sep=',',header=T,stringsAsFactors =FALSE)

#Makes integer values for county codes and state codes:
for(j in 1:nrow(StateCode)){
    StateCode$X1[j] <- as.integer(StateCode$X1[j])
    StateCode$X1.1[j] <- as.integer(StateCode$X1.1[j])
}

#Add state codes to MSA county table:
for(x in 1:nrow(StateCode)){
  for(y in 1:nrow(msaCounty)){
  if(msaCounty$V1[y]==StateCode$AL[x]){
    msaCounty$StateCode[y] = StateCode$X1[x]
  }
  }
}

```

##The first step was to add the conty codes to the cancer file with the associated MSA (metropolitan statistical area) codes:
```{r, echo=FALSE}

pollution$MSACode <-0

#Add the appropriate MSA codes to the pollution table:
for (i in 1:nrow(pollution)) {
  for(j in 1:nrow(msaCounty)){
    if (pollution$County.Code[i] == msaCounty$V6[j] && pollution$State.Code[i] ==msaCounty$StateCode[j]){
      pollution$MSACode[i] <- msaCounty$V4[j]
      
      
    }
  }
}

#Due to issues with the dataset not all of the MSA's from the cancer data set will be in the pollution data set
#The code below counts the number of MSA in the cancer data set that have a corresponding MSA in the pollution data frame:
count =0
for(j in 1:nrow(cancer)){
  for (i in 1:nrow(pollution)) {
    if(is.na(cancer$MSA.Code[j])){
      break
    }
      if (cancer$MSA.Code[j]==pollution$MSACode[i]){
        count = count +1
        cat(count, " ")
        break
      }
    
  
  }
}

#Outputs the new msaCounty county as a .csv (so I can take a break and close R):
write.csv(msaCounty,file = "/Users/Tom/R/Group_Project/processed_tables/msaCountyModified.csv")

#Outputs the new pollution as a .csv:
write.csv(pollution,file = "/Users/Tom/R/Group_Project/processed_tables/pollutionModified.csv")

```


##The same process was repeated for data tables with individual pollutants. This was Ozone:
```{r}

msaCountyModified<- read.csv(file='/Users/Tom/R/Group_Project/processed_tables/msaCountyModified.csv',sep=',',header=T)
Ozone <- read.csv(file='/Users/Tom/R/Group_Project/Data/pollutant/ozone_2010.csv',sep=',',header=T)

#Adds a row for MSA Codes of the Ozone table:
Ozone$MSACode <-0

#Lets add MSA Codes to the Ozone table:
for (i in 1:nrow(Ozone)) {
    if(i%%6000 == 0){           #Adds a display progress indicator
      cat((i/nrow(Ozone))*100,"%     ")
    }
  for(j in 1:nrow(msaCountyModified)){
    if (Ozone$County.Code[i] == msaCountyModified$V6[j] && Ozone$State.Code[i] ==msaCountyModified$StateCode[j]){
      Ozone$MSACode[i] <- msaCountyModified$V4[j]
      
    }
   
  }
}

#Saves as .csv file:
write.csv(Ozone, file = "/Users/Tom/R/Group_Project/Data/pollutant/ozone_2010_Modified.csv")

```

##This was for So2:
```{r}

msaCountyModified<- read.csv(file='/Users/Tom/R/Group_Project/processed_tables/msaCountyModified.csv',sep=',',header=T)
SO2 <- read.csv(file='/Users/Tom/R/Group_Project/Data/pollutant/SO2_2010.csv',sep=',',header=T)

#Add a row for MSA Codes of the SO2 Table:
SO2$MSACode <-0

#Lets add MSA Codes to the SO2 Table:
for (i in 177033:nrow(SO2)) {
    if(i%%6000 == 0){           #Adds a display progress indicator
      cat((i/nrow(SO2))*100,"%     ")
    }
  for(j in 1:nrow(msaCountyModified)){
    if (SO2$County.Code[i] == msaCountyModified$V6[j] && SO2$State.Code[i] ==msaCountyModified$StateCode[j]){
      SO2$MSACode[i] <- msaCountyModified$V4[j]
      
    }
   
  }
}

#Saves as .csv file:
write.csv(SO2, file = "/Users/Tom/R/Group_Project/processed_tables/SO2_2010_Modified.csv")

```

##The next step was to subset the dataframe by MSA code, then to average the AQI for each county. 
##Subsetting the pollution data:
```{r}

#Subset N2O dataset to include only the state code, county code, state name, county name, AQI value, and MSA code:
subN2O <- c("State.Code", "County.Code", "State.Name", "County.Name","AQI","MSACode")
subN2O <- N2O[subN2O]

#Subset CO dataset to include only the state code, county code, state name, county name, AQI value, and MSA code: 
subCO <- c("State.Code", "County.Code", "State.Name", "County.Name", "AQI","MSACode")
subCO <- CO[subCO]

#Subset Ozone dataset to include only the state code, county code, state name, county name, AQI value, and MSA code: 
subOzone <- c("State.Code", "County.Code", "State.Name", "County.Name", "AQI","MSACode")
subOzone <- Ozone[subOzone]

```


##Averaging AQI values:
```{r}

#Average N2O AQI value based on MSA code:
N2Odata <- aggregate(subN2O["AQI"],by=list(subN2O$MSACode),mean,na.rm=TRUE)
#Rename columns:
colnames(N2Odata) <- c("MSACode", "AverageAQI")
#Delete "Other" MSA codes (0 and 99999):
N2Odata <- N2Odata[!(N2Odata$MSACode == 0),]
N2Odata <- N2Odata[!(N2Odata$MSACode == 99999),]
#Save N2Odata as .csv file:
write.csv(N2Odata, file = "/Users/ceacoleman/Desktop/Project/N2Odata.csv")

#Average CO AQI value based on MSA code:
COdata <- aggregate(subCO["AQI"],by=list(subCO$MSACode),mean,na.rm=TRUE)
#Rename columns:
colnames(COdata) <- c("MSACode", "AverageAQI")
#Delete "Other" MSA codes (0 and 99999):
COdata <- COdata[!(COdata$MSACode == 0),]
COdata <- COdata[!(COdata$MSACode == 99999),]
#Save COdata as .csv file:
write.csv(COdata, file = "/Users/ceacoleman/Desktop/Project/COdata.csv")

#Average Ozone AQI value based on MSA code:
Ozonedata <- aggregate(subOzone["AQI"],by=list(subOzone$MSACode),mean,na.rm=TRUE)
#Rename columns:
colnames(Ozonedata) <- c("MSACode", "AverageAQI")
#Delete "Other" MSA codes (0 and 99999):
Ozonedata <- Ozonedata[!(Ozonedata$MSACode == 0),]
Ozonedata <- Ozonedata[!(Ozonedata$MSACode == 99999),]

```

The next step was to subset the ciggerette smoking data by county, and then add the MSA codes:
```{r}

CigSmokers<- read.csv(file='/Users/Tom/R/Group_Project/Data/Health.csv',sep=',',header=T)
myvars <- c("State", "County", "STATECODE", "COUNTYCODE", "County", "Adult.smoking.Value")
CigSmokersProper <- CigSmokers[myvars]

write.csv(CigSmokersProper, file = "/Users/Tom/R/Group_Project/CigSmokersModified")

#Adding MSA Codes to the cig:
myvars <- c("V4", "V6", "StateCode")
MsaCountyClean <- msaCountyModified[myvars]
library(plyr)
MSACountyUse <- rename(MsaCountyClean, c("V4"="MSACODE", "V6"="COUNTYCODE", "StateCode"="STATECODE"))
CigSmokerMSA <- merge(CigSmokersProper,MSACountyUse,by=c("STATECODE","COUNTYCODE")) #Adds MSA Codes to the smoker data
write.csv(CigSmokerMSA, file = "/Users/Tom/R/Group_Project/CigSmokersMSAUSEMEd")

```


##Then this data was used to map the smoking rates, and cancer rates by MSACODE onto the same table (called newTable):
```{r}

install.packages("mapproj")
install.packages("ggmap")
install.packages("DeducerSpatial")

cigSmokersMe <-read.csv(file='/Users/Tom/R/Group_Project/CigSmokersMSAUSEMEd.csv',sep=',',header=T)
myvars <- c("Age.Adjusted.Rate")
cigSmokersMe <- cigSmokersMe[c("Adult.smoking.Value","MSACODE")]
cigSmokersMeat <-cigSmokersMe[!(cigSmokersMe$MSACODE == 99999),] #removes all the 99999 values, because they dont accurately show the location
#Average cigSmokersMeat adult smoking value by MSA code:
cigSmokersMeat <- aggregate(cigSmokersMeat["Adult.smoking.Value"], by=list(cigSmokersMeat$MSACODE), mean, na.rm=TRUE)
#Rename cigSmokersMeat columns:
colnames(cigSmokersMeat) <- c("MSACODE", "Average adult smoking value")

library(plyr)
cancerModified <- rename(cancer, c("MSA.Code"="MSACODE"))
cancerModified <- cancerModified[c("MSACODE", "Age.Adjusted.Rate")]
newTable <- merge(cancerModified, cigSmokersMeat, by=c("MSACODE"))
#newTable contains the cancer rate data, and ciggerate smoking rates, all based on MSA Code. 
#The newTable will be used later on to generate a correlation table
```


##Adding pollutant data into the newTable:
```{r}

#ADD Ozoneto the newTable:
newTable <- merge(newTable, Ozonedata,by=c("MSACODE"))

#ADD NO2 to the newTable:
newTable <- merge(newTable,N2Odata,by=c("MSACODE"))

#ADD CO to the newTable:
newTable <- merge(newTable,COdata,by=c("MSACODE"))
```


##Using newTable to create a correlation table, and then generating the corrplot:
```{r}

library(corrplot)
cortable <- cor(newTable)
colnames(cortable) = c("MSACODE", "Age Adjusted Cancer Rate", "Average Adult Smoking Rate", "N2O AQI", "OZONE AQI", "CO AQI")

jpeg('PollutantCorrelation.jpg')
corrplot(cortable, method = "number")
dev.off()

```

##In conclusion:
```{r}
  There is a strong positive correlation between ciggeratte smoking and Age.Adjusted Cancer Rate.
  There is a strong positive correlation between the presence of one pollutant and the presence of others.
  The above two points are in line with what was expected.
  A higher AQI means the air quality is worse. It was unexpected that a higher AQI would be negatively correlated with age adjusted cancer rate.

```

